---
title: "Introduction to enrichit"
format: 
    html:
        theme: none
        embed-resources: true
        fontsize: 1.1em
        linestretch: 1.7
author: "Guangchuang Yu\\

        School of Basic Medical Sciences, Southern Medical University"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{enrichit}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


# Introduction

Functional enrichment analysis is a staple in bioinformatics for interpreting lists of genes identified from omics experiments. `enrichit` provides fast, C++-based implementations of three of the most widely used methods:

1.  **Over-Representation Analysis (ORA)**
2.  **Gene Set Enrichment Analysis (GSEA)**
3.  **Network-based Set Enrichment Analysis (NSEA)**

The package is designed to be efficient and easy to integrate into existing workflows, with a focus on performance and standardized output formats.

# Installation

You can install `enrichit` from GitHub:

```{r eval=FALSE}
devtools::install_github("YuLab-SMU/enrichit")
```

# Over-Representation Analysis (ORA)

ORA determines whether a set of genes of interest (e.g., differentially expressed genes) is enriched in a known gene set (e.g., a biological pathway) more than would be expected by chance.

## Method

`enrichit` implements ORA using the **hypergeometric distribution** (one-sided Fisher's exact test). The p-value is calculated as the probability of observing at least $k$ genes from the specific gene set in the selected list of $n$ genes, given a background population (universe) of $N$ genes containing $M$ genes from that set.

$$
p = 1 - \sum_{i=0}^{k-1} \frac{\binom{M}{i} \binom{N-M}{n-i}}{\binom{N}{n}}
$$

## Example

```{r}
library(enrichit)

# Simulate a universe of 1000 genes
universe <- paste0("Gene", 1:1000)

# Define gene sets
gene_sets <- list(
  PathwayA = paste0("Gene", 1:50),       # Genes 1-50
  PathwayB = paste0("Gene", 800:850)     # Genes 800-850
)

# Select 'significant' genes (e.g., top 20 genes)
# PathwayA should be enriched
sig_genes <- paste0("Gene", 1:20)

# Run ORA
ora_result <- ora(
  gene = sig_genes,
  gene_sets = gene_sets,
  universe = universe
)

# View results
as.data.frame(ora_result)
```

# Gene Set Enrichment Analysis (GSEA)

GSEA evaluates whether a defined set of genes shows statistically significant, concordant differences between two biological states. Unlike ORA, GSEA uses the entire ranked list of genes, avoiding the need for arbitrary thresholds to select "significant" genes.

## Method

`enrichit` offers a fast C++ implementation of GSEA. It calculates an **Enrichment Score (ES)** that reflects the degree to which a gene set is over-represented at the top or bottom of a ranked list of genes.

The package supports different methods for p-value calculation:

1.  **Multilevel (`method = "multilevel"`)**: This is the default and recommended method. It uses an adaptive multi-level splitting Monte Carlo approach to estimate low p-values efficiently with high accuracy, similar to the `fgsea` package.
2.  **Simple Permutation (`method = "permute"`)**: Standard permutation of gene labels.
3.  **Sample Permutation (`method = "sample"`)**: Random sampling of gene sets (faster but less rigorous for some null hypotheses).

## Example

```{r}
# Generate synthetic ranked gene list
set.seed(42)
geneList <- sort(rnorm(1000), decreasing = TRUE)
names(geneList) <- paste0("Gene", 1:1000)

# Define gene sets
# PathwayTop is enriched at the top (positive ES)
# PathwayBottom is enriched at the bottom (negative ES)
gene_sets <- list(
  PathwayTop = names(geneList)[1:50],
  PathwayBottom = names(geneList)[951:1000],
  PathwayRandom = sample(names(geneList), 50)
)

# Run GSEA using the multilevel method
gsea_result <- gsea(
  geneList = geneList,
  gene_sets = gene_sets,
  method = "multilevel",
  nPerm = 1000,    # Base permutations
  minGSSize = 10,
  maxGSSize = 500
)

# View results
head(gsea_result)
```

# Working with GSON

`enrichit` works seamlessly with `GSON` objects, which are used to store gene set information along with metadata. The `GSON` class is defined in the `r yulab.utils::CRANpkg("gson")` package. It provides a structured way to handle gene sets, including gene identifiers, gene set names, and other associated information.

```{r eval=FALSE}
# Assuming you have a GSON object 'g'
# result <- gsea_gson(geneList = geneList, gson = g)
```

# Network-based Set Enrichment Analysis (NSEA)

Traditional enrichment methods treat pathways as independent "bags of genes," ignoring the biological interactions between them. NSEA addresses this by leveraging a biological network (e.g., a Protein-Protein Interaction network) to propagate the initial gene signals before performing enrichment. 

## Method

`enrichit` implements NSEA by combining a highly optimized **Random Walk with Restart (RWR)** algorithm with the fast multilevel GSEA engine. 

1. **Propagation**: The initial gene scores are mapped onto the network as seed probabilities. An extremely fast C++ RWR algorithm (powered by `RcppEigen` sparse matrix multiplication) diffuses these probabilities until a stationary distribution is reached.
2. **Ranking**: Genes are ranked based on their final network propagation scores, which reflect their topological proximity to the initial gene list.
3. **Enrichment**: The ranked list is directly passed to the internal GSEA engine. Because propagation scores are non-negative, the enrichment is evaluated in a one-sided manner (`scoreType = "pos"`).

## Example

```{r}
library(enrichit)
library(Matrix)

# 1. Create a mock network (edge list)
set.seed(123)
edges <- data.frame(
    from = sample(LETTERS[1:10], 20, replace = TRUE),
    to = sample(LETTERS[1:10], 20, replace = TRUE),
    weight = runif(20)
)
edges <- edges[edges$from != edges$to, ] # remove self loops

# 2. Create a mock geneList with evidence scores
geneList <- setNames(runif(5), sample(LETTERS[1:10], 5))

# 3. Define gene sets
gene_sets <- list(
    PathwayA = c("A", "B", "C", "D"),
    PathwayB = c("E", "F", "G", "H"),
    PathwayC = c("I", "J", "A")
)

# 4. Run NSEA
nsea_result <- nsea(
    geneList = geneList,
    network = edges,
    gene_sets = gene_sets,
    p = 0.5,         # restart probability
    minGSSize = 2,
    maxGSSize = 10
)

# View results
head(nsea_result)
```

If you plan to run multiple gene lists on the same network, you can significantly speed up the process by preparing the network once:

```{r eval=FALSE}
net <- prepare_network(edges)
res1 <- nsea(geneList = list1, network = net, gene_sets = gene_sets)
res2 <- nsea(geneList = list2, network = net, gene_sets = gene_sets)
```

# Bayesian Over-Representation Analysis

For ORA results, `enrichit` also provides a Bayesian framework (`bayes_enrich()`) to estimate the posterior probability of explanatory terms, helping to select active pathways while penalizing redundancy.

```{r eval=FALSE}
# After obtaining an ORA result object
# bayes_res <- bayes_enrich(ora_result)
# summary <- bayes_summary(bayes_res)
```

# Session Info

```{r}
sessionInfo()
```
